---
title: "Mock communtiy quantification"
author: "Jamie McMurtrie"
date: "07/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/jm1106/malawi_tilapia")
```

##Introduction
We used the ZymoBIOMICS® Microbial Community DNA standard, lot number ZRC190811.
expected ratios found here https://files.zymoresearch.com/protocols/_d6305_d6306_zymobiomics_microbial_community_dna_standard.pdf

Plot a relative abundance bar chart of expected vs sequenced controls. e.g.
https://github.com/yanxianl/Li_AqFl1-Microbiota_2021/blob/master/code/08_taxonomy.Rmd
https://github.com/yanxianl/Li_AqFl1-Microbiota_2021/blob/master/data/reference/mock_expected.tsv


```{r libraries, include=TRUE}
library(here); packageVersion("here") # make the relative paths work
library(tidyverse); packageVersion("tidyverse")
library(phyloseq); packageVersion("phyloseq")
library(cowplot); packageVersion("cowplot") # for density plots
```
```{r phyloseq tidy up functions}
tidyPS <- function(ps, RA){
  ps <- prune_taxa(taxa_sums(ps) > 0, ps)
  ps <- prune_samples(sample_sums(ps) > 0, ps)
  if(RA == TRUE){
    ps <- transform_sample_counts(ps, function(x){x / sum(x)})
  }
  return(ps)} 
```
## Read in files
```{r libraries, include=TRUE}
mock_expected <- read.csv(here("data", "references", "mock_community_expected.csv"))

## Load data ####
## Combine ASV count table and taxonomy table from V4_DADA2_generate_ASV_table.R
## script, plus metadata table, into a phyloseq object
## Load metadata, with first column (Sample) as row names to match ASV table
#metadata <- read.table(here("data/raw/malawi_tilapia_metadata.txt"), header = TRUE,
#                       row.names = 1)
#seqtab <- readRDS(here("data/process/V4_ASVtable_final.rds"))
#tax <- readRDS(here("data/process/V4_tax_bayesian_final.rds"))
#
#
## Combine into phyloseq object
#ps <- phyloseq(otu_table(seqtab, taxa_are_rows=FALSE), 
#               sample_data(metadata), 
#               tax_table(tax))
#
## Remove large temporary objects
#rm(seqtab); rm(tax)
#ps.cont <- subset_samples(ps, sample_type %in% "control")
#mock_names <- grep("mock", sample_names(ps.cont), value = TRUE)
#ps.ra.mock <- subset_samples(ps.cont, sample_names(ps.cont) %in% mock_names) %>% tidyPS(TRUE)

ps_post_decontam <- readRDS(here("data/process/V4_phyloseq_obj_after_decontam.rds"))
ps_post_decontam %>% sample_data()

ps_mocks <- subset_samples(ps_post_decontam, sample_type_detailed %in% c("control_mock_DNA", "control_mock_extr")) %>% tidyPS(TRUE)
#ps_mocks_2 <- subset_samples(ps_post_decontam, sample_type_detailed %in% "control_mock_DNA") %>% tidyPS(TRUE)

#change to ASV names
seqs <- Biostrings::DNAStringSet(taxa_names(ps_mocks))
names(seqs) <- taxa_names(ps_mocks)
ps_mocks_2 <- merge_phyloseq(ps_mocks, seqs)
taxa_names(ps_mocks_2) <- paste0("ASV", seq(ntaxa(ps_mocks_2)))
View(ps_mocks_2@tax_table@.Data)

#merge at genus level (note 2x bacillus ASVs), we will evaluate all at genus level, only a couple at species definition (not that that can really be trusted)
ps_mocks_gen <- tax_glom(ps_mocks_2, "Genus", NArm = FALSE)
tax_table(ps_mocks_gen)

#expec_genus <- mock_expected[,"Taxonomy"] %>% str_split(" ")
#expec_genus[1] 
#seperate(mock_expected, )
#separate(mock_expected, "Taxonomy", into = NA, sep = " ")
```


check listeria in pre-decontam object
```{r}
mock_expected <- read.csv(here("data", "references", "mock_community_expected.csv"))

metadata <- read.table(here("data/raw/malawi_tilapia_metadata.txt"), header = TRUE,
                       row.names = 1)
seqtab <- readRDS(here("data/process/V4_ASVtable_final.rds"))
tax <- readRDS(here("data/process/V4_tax_bayesian_final.rds"))


# Combine into phyloseq object
ps <- phyloseq(otu_table(seqtab, taxa_are_rows=FALSE), 
               sample_data(metadata), 
               tax_table(tax))

ps_full_mocks <- subset_samples(ps, sample_type_detailed %in% c("control_mock_DNA", "control_mock_extr")) %>% tidyPS(TRUE)

seqs <- Biostrings::DNAStringSet(taxa_names(ps_full_mocks))
names(seqs) <- taxa_names(ps_full_mocks)
ps_full_mocks <- merge_phyloseq(ps_full_mocks, seqs)
taxa_names(ps_full_mocks) <- paste0("ASV", seq(ntaxa(ps_full_mocks)))
View(ps_full_mocks@tax_table@.Data)

#merge at genus level (note 2x bacillus ASVs), we will evaluate all at genus level, only a couple at species definition (not that that can really be trusted)
ps_mocks_gen <- tax_glom(ps_full_mocks, "Genus", NArm = FALSE)
tax_table(ps_mocks_gen)

```

For this step I need to add a new sample to this object as the expected, and provide expected values for each equivilant ASV.
Also need to group the remaining non-expected ASVs into an 'other'. Get a vector of the expecetd genus, then use this to select the other genera to merge under phyloseq merge_taxa.
```{r format expected data, include=TRUE, echo=FALSE}
mock_taxonomy <- mock_expected %>% separate(Taxonomy, c("Genus", "Species"))
expec_genus <- mock_taxonomy$Genus
mock_taxa <- ps_mocks_gen %>% subset_taxa(Genus %in% expec_genus) %>% taxa_names()
mock_taxa <- c(mock_taxa, "ASV4")
#for some reasons the e coli from the mock community has been assignedby silva as "Escherichia/Shigella" so we shall mannual add the corresponding ASV4
all_taxa <- taxa_names(ps_mocks_gen)
taxa_to_merge <- setdiff(all_taxa, mock_taxa)

ps_mock_megre <- merge_taxa(ps_mocks_gen, taxa_to_merge)
taxa_names(ps_mock_megre)

ps_mock_megre
```

```{r construct expected sampled}
mock_exp_sam <- data.frame(NA,NA,NA,NA,"control_expected", "control_expected", NA, "sample", NA, NA,NA, FALSE)
names(mock_exp_sam) <- colnames(sample_data(ps_mock_megre))
rownames(mock_exp_sam) <- "Mock_expected"

mock_exp_otu <- data.frame(0.174,0.155,0.101,0.184,0.104,0.141,0.099,0.042,0)
names(mock_exp_otu) <- colnames(otu_table(ps_mock_megre))
rownames(mock_exp_otu) <- "Mock_expected"

new_sample_data <- rbind(sample_data(ps_mock_megre), mock_exp_sam)
new_otu_tab <- rbind(otu_table(ps_mock_megre), mock_exp_otu)

final_phyloseq <- phyloseq(sample_data(new_sample_data),
                           otu_table(new_otu_tab, taxa_are_rows = FALSE),
                           tax_table(ps_mock_megre@tax_table@.Data))
final_phyloseq


desired_order <- sample_names(final_phyloseq) #for some reason the samples in plot bar get reordered away from those provided.
theme_set(theme_minimal()) 
#plot_bar(final_phyloseq, fill = "Genus", facet_grid = ~sample_type)
colour_pal <- c("#332288", "#88CCEE", "#44AA99", "#117733", "#999933", "#DDCC77", "#CC6677", "#882255", "#C5C5C5")
p <- plot_bar(final_phyloseq, fill = "Genus") + scale_fill_manual(values = colour_pal)
p$data$Sample <- factor(p$data$Sample, levels = desired_order)
p

plot_bar_2 <-  function (physeq, x = "Sample", y = "Abundance", fill = NULL, title = NULL, facet_grid = NULL, border_color = NA) 
{
  mdf = psmelt(physeq)
  p = ggplot(mdf, aes_string(x = x, y = y, fill = fill))
  p = p + geom_bar(stat = "identity", position = "stack",  color = border_color)
  p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
  if (!is.null(facet_grid)) {
    p <- p + facet_grid(facet_grid)
  }
  if (!is.null(title)) {
    p <- p + ggtitle(title)
  }
  return(p)
}

p <- plot_bar_2(final_phyloseq, fill = "Genus") + geom_bar(stat = "identity") + scale_fill_manual(values = colour_pal, na.value = "#C5C5C5")
p$data$Sample <- factor(p$data$Sample, levels = desired_order)
p

ggsave(here::here("results", "figures", "mock_barplot_16s.svg"), plot = p, width = 10, height = 8, dpi = 600)
```

## Now 18S

```{r}
#mock_expected <- read.csv(here("data", "references", "mock_community_expected.csv"))

metadata <- read.table(here("data/raw/malawi_tilapia_metadata.txt"), header = TRUE,
                       row.names = 1)
seqtab <- readRDS(here("data/process/V9_seqtab_nochim.rds"))
tax <- readRDS(here("data/process/V9_tax_final_silva.rds"))


# Combine into phyloseq object
ps <- phyloseq(otu_table(seqtab, taxa_are_rows=FALSE), 
               sample_data(metadata), 
               tax_table(tax))

ps_full_mocks <- ps %>% subset_taxa(Kingdom == "Eukaryota") %>% 
                  subset_samples(sample_type_detailed %in% c("control_mock_DNA", "control_mock_extr")) %>% tidyPS(TRUE) 

seqs <- Biostrings::DNAStringSet(taxa_names(ps_full_mocks))
names(seqs) <- taxa_names(ps_full_mocks)
ps_full_mocks <- merge_phyloseq(ps_full_mocks, seqs)
taxa_names(ps_full_mocks) <- paste0("ASV", seq(ntaxa(ps_full_mocks)))
View(ps_full_mocks@tax_table@.Data)

#merge at family level (note 2x Cryptococcaceae ASVs), we will evaluate all at family level, only a couple at genus definition (not that that can really be trusted)
ps_mocks_gen <- tax_glom(ps_full_mocks, "Family", NArm = FALSE)
tax_table(ps_mocks_gen)

```


```{r format expected data, include=TRUE, echo=FALSE}
expec_family <- c("Saccharomycetaceae", "Cryptococcaceae")
mock_taxa <- ps_mocks_gen %>% subset_taxa(Family %in% expec_family) %>% taxa_names()

all_taxa <- taxa_names(ps_mocks_gen)
taxa_to_merge <- setdiff(all_taxa, mock_taxa)

ps_mock_megre <- merge_taxa(ps_mocks_gen, taxa_to_merge)
taxa_names(ps_mock_megre)

ps_mock_megre

mock_exp_sam <- data.frame(NA,NA,NA,NA,"control_expected", "control_expected", NA, "sample", NA, NA,NA, FALSE)
names(mock_exp_sam) <- colnames(sample_data(ps_mock_megre))
rownames(mock_exp_sam) <- "Mock_expected"

mock_exp_otu <- data.frame(0,0.5,0.5)
names(mock_exp_otu) <- colnames(otu_table(ps_mock_megre))
rownames(mock_exp_otu) <- "Mock_expected"

new_sample_data <- rbind(sample_data(ps_mock_megre), mock_exp_sam)
new_otu_tab <- rbind(otu_table(ps_mock_megre), mock_exp_otu)

final_phyloseq <- phyloseq(sample_data(new_sample_data),
                           otu_table(new_otu_tab, taxa_are_rows = FALSE),
                           tax_table(ps_mock_megre@tax_table@.Data))
final_phyloseq

desired_order <- sample_names(final_phyloseq) #for some reason the samples in plot bar get reordered away from those provided.

p2 <- plot_bar_2(final_phyloseq, fill = "Family") + geom_bar(stat = "identity") + scale_fill_manual(values = colour_pal, na.value = "#C5C5C5")
p2$data$Sample <- factor(p2$data$Sample, levels = desired_order)
p2

ggsave(here::here("results", "figures", "mock_barplot_18s.svg"), plot = p2, width = 10, height = 8, dpi = 600)
```

