---
title: "Tilapia species differences"
author: "Jamie McMurtrie"
date: "21/6/2021"
output: html_document

---
# Supplementary figure 5
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/jm1106/Malawi_Tilapia_Microbiomes")
```

## R Setup Libraries

```{r libraries, include=TRUE, echo=TRUE}
library(tidyverse); packageVersion("tidyverse")
library(phyloseq); packageVersion("phyloseq")
library(RColorBrewer); packageVersion("RColorBrewer")
library(ggpubr); packageVersion("ggpubr") # for publication-quality ggplots
library(cowplot); packageVersion("cowplot") # for density plots
library(here); packageVersion("here") # make the relative paths work
library(DivNet); packageVersion("DivNet")
library(breakaway); packageVersion("breakaway")
library(factoextra); packageVersion("factoextra") # for lovely PCA biplots
```


```{r seed and plot settings, include=TRUE, echo=TRUE}
theme_set(theme_bw())
set.seed(1234)

theme_set(theme(panel.background = element_blank(), axis.line = element_line(colour = "black")))
#when plotting with ggplot, removes grid lines etc for a clean plot 
species_colour <- c("#ffabaf", "#8a2424", "#d63638")
```

A useful function for cleaning up phyloseq objects after subsetting to remove any zero taxa or samples, and optionally transform counts to relative abundance. Second function for removing undesired taxa (by ASV name) from phyloseq objects.
```{r phyloseq tidy up functions}
tidyPS <- function(ps, RA){
  ps <- prune_taxa(taxa_sums(ps) > 0, ps)
  ps <- prune_samples(sample_sums(ps) > 0, ps)
  if(RA == TRUE){
    ps <- transform_sample_counts(ps, function(x){x / sum(x)})
  }
  return(ps)} 
```

Reading in data and preparing phyloseq objects for analysis.
```{r data setup}
s16_ps <- readRDS(here("data", "process", "V4_phyloseq_obj_with_tree_prev_filt.rds")) #can't start variable names with numbers, hence s16 / s18 herein refers to 16s 18s

s16_ps_swab <- subset_samples(s16_ps, sample_type %in% "Swab") %>% tidyPS(., FALSE) 

s16_ps_swab_og <- subset_samples(s16_ps, sample_type %in% "Swab") %>% tidyPS(., FALSE) #the original non rarefied object is used for beta diversity (compositional log ratio approaches)

set.seed(1234)
s16_ps_swab <- rarefy_even_depth(s16_ps_swab) #removes 492 ASVs
#the rarefied object is used for alpha diversity calcs

sample_data(s16_ps_swab)$Pond <- as.factor(sample_data(s16_ps_swab)$Pond) #this changes time points from a continuous scale to categoric
```

## All species

```{r all species alpha diversity}
sample_data(s16_ps_swab)

species <- c(rep.int("Coptodon rendalli",11), rep.int("Tilapia sparmanii", 3), "Coptodon rendalli", "Tilapia sparmanii", "Coptodon rendalli", "Tilapia sparmanii", rep.int("Coptodon rendalli",5), rep.int("Oreochromis shiranus",9))

sample_data(s16_ps_swab)$Species <- species #add species labels to each sample

wunifrac_dist <- phyloseq::distance(s16_ps_swab, method="unifrac", weighted=F)
ordination <- ordinate(s16_ps_swab, method="PCoA", distance=wunifrac_dist)
plot_ordination(s16_ps_swab, ordination, color = "Species") + theme(aspect.ratio=1)

#plot_richness(s16_ps_swab, x="Species", measures = c("Chao1", "Shannon"))

alpha_ests <- estimate_richness(s16_ps_swab, measures = c("Chao1", "Shannon")) %>%
              cbind(sample_data(s16_ps_swab)[,"Species"]) %>%
              cbind(sample_data(s16_ps_swab)[,"Pond"])

chao1_wlech <- oneway.test(alpha_ests$Chao1 ~ alpha_ests$Species)
chao1_wlch_p <- paste("Welch's ANOVA, p = ", round(chao1_wlech$p.value, digits = 3))
  
chao1_p <- ggplot(data = alpha_ests, aes(x = Species, y = Chao1, color = Species)) +
              geom_point(shape=19, alpha=0.4, stroke = 1, size = 3, aes(color=Species), show.legend=F) +
              stat_summary(fun = mean, geom = "crossbar", width = 0.8,
                       aes(color=Species), show.legend = F) +
            scale_color_manual(values = species_colour) +
            theme(axis.text.x = element_text(angle = 45, vjust = 0.5))+ #, vjust = 0.5, hjust=1)) +
            annotate("text", x = 2, y = 800, label = chao1_wlch_p)

shan_wlech <- oneway.test(alpha_ests$Shannon ~ alpha_ests$Species)
shan_wlch_p <- paste("Welch's ANOVA, p = ", round(shan_wlech$p.value, digits = 3))

shan_p <- ggplot(data = alpha_ests, aes(x = Species, y = Shannon, color = Species)) +
              geom_point(shape=19, alpha=0.4, stroke = 1, size = 3, aes(color=Species), show.legend=F) +
              stat_summary(fun = mean, geom = "crossbar", width = 0.8,
                       aes(color=Species), show.legend = F) +
           scale_color_manual(values = species_colour) +
            theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
          annotate("text", x = 2, y = 6, label = shan_wlch_p)

chao1_p
shan_p
```

```{r all species clr beta diversity}
samples_V4 <- as.data.frame(sample_data(s16_ps_swab))
samples_V4 <- as_tibble(samples_V4, rownames = "SampleID")

V4_czm <- zCompositions::cmultRepl(otu_table(s16_ps_swab_og), method = "CZM") #note using non-rarefied object

V4_comp <- CoDaSeq::codaSeq.clr(V4_czm) 

# Run the PCA on the CLR-transformed data
V4_PCA <- prcomp(V4_comp)

pind_V4 <- fviz_pca_ind(V4_PCA, 
                            geom.ind = "point", # show points only (but not "text"))
                            pointsize = 4,
                            pointshape = 19, alpha = 0.7,
                            col.ind =  samples_V4$Species, # Grouping variable
                            #shape.ind = df$Time_point,
                            palette = species_colour,
                            #addEllipses = TRUE, 
                            legend.title = "Culture species",
                            invisible="quali") #to remove group centrodi large points
pind_V4


region <- c(rep("Blantyre",23), rep("Maldeco",9))
samples_V4$Region <- region

  
clr_dist <- dist(V4_comp, method = "euclidean") #calculate distance
dispr <- vegan::betadisper(clr_dist, samples_V4$Species)
print(dispr)
dispr_reg <- vegan::betadisper(clr_dist, samples_V4$Region)
print(dispr_reg)
print(vegan::permutest(dispr, pairwise = TRUE))

adonis_species <- vegan::adonis(clr_dist ~ samples_V4$Species)
print(adonis_species)
adonis_region <- vegan::adonis(clr_dist ~ samples_V4$Region)
print(adonis_region)

```

COmbine figures
```{r combine}
alpha_plots <- cowplot::plot_grid(chao1_p, shan_p, ncol = 2)
alpha_beta_plots <- cowplot::plot_grid(alpha_plots, pind_V4, nrow = 2)
alpha_beta_plots
ggsave(here("results", "figures", "species_2.svg"), plot = alpha_beta_plots, width = 8, height = 8, dpi = 600)
```